---
title: "Assignment 5: Effect of ad-variety and session-level history on customer response to mobile advertising"
author: "Tondreka Robles"
date: "Spring B 2026, MKT 472"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
set.seed(472)
```

```{r libraries-and-data}
library(rpart)
library(rpart.plot)
library(xgboost)
library(knitr)

load("data/variety_train.RData")
load("data/variety_test.RData")

within_vars <- c("variety", "rep", "adimpsession")
pre_vars <- c("imptotal", "ctruser", "varietytotal", "adimptotal")
full_vars <- c("timeofday", "imptotal", "ctruser", "varietytotal",
               "adimptotal", "variety", "rep", "adimpsession")
```

## 1. Descriptive analysis

### 1(a)

```{r q1a}
observed_ctr <- mean(variety_train$click)
average_past_ctr <- mean(variety_train$ctruser)

kable(data.frame(
  Measure = c("Observed CTR", "Average user's past CTR"),
  Value = c(observed_ctr, average_past_ctr)
), digits = 4)
```

The observed CTR is 0.1134, and the average of users' past CTR is 0.1165. These
numbers are close, which makes sense because users with higher past click rates
are also more likely to click in the current session. They are not exactly the
same because the current data is for one specific EA ad impression.

### 1(b)

```{r q1b}
hist(variety_train$variety,
     main = "Histogram of In-Session Variety",
     xlab = "Variety",
     col = "lightblue")

hist(variety_train$varietytotal,
     main = "Histogram of Pre-Session Variety",
     xlab = "Variety Total",
     col = "lightgreen")
```

In-session variety is limited because it only comes from the first seven ads in
the session. Most users saw around 3 or 4 different ads. Pre-session variety has
a wider spread because it is based on the user's entire past history before the
session.

### 1(c)

```{r q1c}
cor.test(variety_train$variety, variety_train$rep)
```

The correlation is about -0.70, which is a strong negative relationship. This is
expected because if a user sees more different ads, there is less repetition.

### 1(d)

```{r q1d}
avg_ctr <- tapply(variety_train$click, variety_train$variety, mean)

plot(as.numeric(names(avg_ctr)), avg_ctr,
     type = "b",
     xlab = "In-session variety",
     ylab = "Average CTR",
     main = "Average CTR by In-Session Variety",
     col = "blue",
     pch = 19)

avg_ctr
```

CTR increases as in-session variety increases. Users are less likely to click
when they saw only one or two different ads earlier in the session, and more
likely to click when they saw more variety.

### 1(e)

This effect can be interpreted as causal because the ads shown before the eighth
impression were randomly assigned. Since the variety was created by the
experiment, the change in CTR is not only a regular correlation.

## 2. Within-session models

### 2(a) and 2(b): CART model

```{r q2-cart, fig.width=8, fig.height=5.5}
cart_within <- rpart(click ~ variety + rep + adimpsession,
                     data = variety_train,
                     cp = 0.00032)

rpart.plot(cart_within)

sum(cart_within$frame$var == "<leaf>")
cart_within$variable.importance
```

This tree has 5 leaves. It splits mainly on `variety`, so in-session variety is
the most important variable. The tree shows higher predicted CTR for higher
levels of variety. `rep` and `adimpsession` are omitted from the final splits.

### 2(c): CART prediction

```{r q2-cart-pred}
variety_test$withinsession.CART.pred <- predict(cart_within, newdata = variety_test)
head(variety_test$withinsession.CART.pred)
```

### 2(d) and 2(e): XGBoost model and prediction

```{r q2-xgb}
xgb_within <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    eta = 0.1,
    max_depth = 6,
    subsample = 1,
    colsample_bytree = 1,
    min_child_weight = 5,
    gamma = 5
  ),
  data = xgb.DMatrix(as.matrix(variety_train[, within_vars]), label = variety_train$click),
  nrounds = 100,
  verbose = 0
)

variety_test$withinsession.xgb.pred <- predict(
  xgb_within,
  xgb.DMatrix(as.matrix(variety_test[, within_vars]))
)

head(variety_test$withinsession.xgb.pred)
```

## 3. Pre-session models

### 3(a) and 3(b): CART model

```{r q3-cart, fig.width=8, fig.height=5.5}
cart_pre <- rpart(click ~ imptotal + ctruser + varietytotal + adimptotal,
                  data = variety_train,
                  cp = 0.00032)

rpart.plot(cart_pre)

sum(cart_pre$frame$var == "<leaf>")
cart_pre$variable.importance
```

This tree has 7 leaves. It splits mainly on `ctruser`, meaning the user's past
CTR is the most important pre-session predictor. Users with higher past CTR have
higher predicted click rates. Other pre-session variables are less important or
omitted.

### 3(c): CART prediction

```{r q3-cart-pred}
variety_test$presession.CART.pred <- predict(cart_pre, newdata = variety_test)
head(variety_test$presession.CART.pred)
```

### 3(d) and 3(e): XGBoost model and prediction

```{r q3-xgb}
xgb_pre <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    eta = 0.1,
    max_depth = 6,
    subsample = 1,
    colsample_bytree = 1,
    min_child_weight = 5,
    gamma = 5
  ),
  data = xgb.DMatrix(as.matrix(variety_train[, pre_vars]), label = variety_train$click),
  nrounds = 100,
  verbose = 0
)

variety_test$presession.xgb.pred <- predict(
  xgb_pre,
  xgb.DMatrix(as.matrix(variety_test[, pre_vars]))
)

head(variety_test$presession.xgb.pred)
```

## 4. Full models

### 4(a) and 4(b): CART model

```{r q4-cart, fig.width=9, fig.height=7}
cart_full <- rpart(click ~ timeofday + imptotal + ctruser + varietytotal +
                     adimptotal + variety + rep + adimpsession,
                   data = variety_train,
                   cp = 0.00032)

rpart.plot(cart_full)

sum(cart_full$frame$var == "<leaf>")
cart_full$variable.importance
```

The full CART model has 17 leaves. It uses both past behavior and within-session
behavior. The most important variables are `ctruser` and `variety`. This means
past CTR is important, but in-session variety still adds useful information.

### 4(c): CART prediction

```{r q4-cart-pred}
variety_test$full.CART.pred <- predict(cart_full, newdata = variety_test)
head(variety_test$full.CART.pred)
```

### 4(d) and 4(e): XGBoost model and prediction

```{r q4-xgb}
xgb_full <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    eta = 0.1,
    max_depth = 4,
    subsample = 1,
    colsample_bytree = 1,
    min_child_weight = 5,
    gamma = 5
  ),
  data = xgb.DMatrix(as.matrix(variety_train[, full_vars]), label = variety_train$click),
  nrounds = 100,
  verbose = 0
)

variety_test$full.xgb.pred <- predict(
  xgb_full,
  xgb.DMatrix(as.matrix(variety_test[, full_vars]))
)

head(variety_test$full.xgb.pred)
```

## 5. Model evaluation

```{r q5-functions}
auc_calc <- function(actual, pred) {
  r <- rank(pred, ties.method = "average")
  n1 <- sum(actual == 1)
  n0 <- sum(actual == 0)
  (sum(r[actual == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

rig_calc <- function(actual, pred) {
  pred <- pmin(pmax(pred, 1e-15), 1 - 1e-15)
  base <- mean(actual)
  model_loss <- -mean(actual * log2(pred) + (1 - actual) * log2(1 - pred))
  base_loss <- -mean(actual * log2(base) + (1 - actual) * log2(1 - base))
  1 - model_loss / base_loss
}

model_columns <- c("withinsession.CART.pred", "withinsession.xgb.pred",
                   "presession.CART.pred", "presession.xgb.pred",
                   "full.CART.pred", "full.xgb.pred")

model_names <- c("Within CART", "Within XGBoost",
                 "Pre-session CART", "Pre-session XGBoost",
                 "Full CART", "Full XGBoost")

performance <- data.frame(
  Model = model_names,
  AUC = sapply(model_columns, function(x) auc_calc(variety_test$click, variety_test[[x]])),
  RIG = sapply(model_columns, function(x) rig_calc(variety_test$click, variety_test[[x]]))
)

kable(performance, digits = 4)
```

The best model is the full XGBoost model because it has the highest AUC and RIG.
The pre-session models perform better than the within-session models alone, but
the full model performs best because it combines both sources of information.

## 6. Main findings

Pre-session history is more valuable than within-session history by itself,
especially because `ctruser` is a strong predictor. However, within-session
history still matters. The best results come from using both pre-session and
within-session variables together.

Within-session variety has a positive effect on click behavior. Users who saw a
higher variety of ads earlier in the session were more likely to click the EA ad.

## 7. Business implications

### 7(a) and 7(b)

```{r q7-top5000}
best_model <- "full.xgb.pred"

top5000 <- variety_test[order(variety_test[[best_model]], decreasing = TRUE), ][1:5000, ]

kable(data.frame(
  Measure = c("Actual CTR for top 5000", "Average predicted CTR for top 5000"),
  Value = c(mean(top5000$click), mean(top5000[[best_model]]))
), digits = 4)
```

The actual CTR for the top 5000 impressions is 0.1954, and the average predicted
CTR is 0.1872. These are fairly close, so the model does a reasonable job of
identifying high-value impressions.

### 7(c)

```{r q7-roi}
cost_per_impression <- 0.05
value_per_click <- 2

baseline_spend <- nrow(variety_test) * cost_per_impression
baseline_gain <- sum(variety_test$click) * value_per_click
baseline_roi <- (baseline_gain - baseline_spend) / baseline_spend

top_spend <- nrow(top5000) * cost_per_impression
top_gain <- sum(top5000$click) * value_per_click
top_roi <- (top_gain - top_spend) / top_spend

kable(data.frame(
  Strategy = c("Buy all impressions", "Buy top 5000 impressions"),
  Spend = c(baseline_spend, top_spend),
  Gain = c(baseline_gain, top_gain),
  ROI = c(baseline_roi, top_roi)
), digits = 4)
```

Buying all impressions gives an ROI of 3.5053. Buying only the top 5000
impressions gives an ROI of 6.8160, which is much better.

### 7(d)

```{r q7-allocation}
ordered_test <- variety_test[order(variety_test[[best_model]], decreasing = TRUE), ]

allocation <- data.frame(impressions = seq(500, 20000, by = 500))
allocation$clicks <- sapply(allocation$impressions, function(n) sum(ordered_test$click[1:n]))
allocation$ad_revenue <- allocation$clicks * value_per_click
allocation$ad_cost <- allocation$impressions * cost_per_impression
allocation$ad_profit <- allocation$ad_revenue - allocation$ad_cost
allocation$price_promo_spend <- 1000 - allocation$ad_cost
allocation$price_promo_profit <- allocation$price_promo_spend * 5
allocation$total_profit <- allocation$ad_profit + allocation$price_promo_profit

best_allocation <- allocation[which.max(allocation$total_profit), ]

kable(best_allocation, digits = 4)
```

The best allocation is to buy the top 5,500 ad impressions. This costs $275 and
produces $2,120 in advertising revenue. EA should spend the remaining $725 on
price promotions. This gives the highest total profit among the options checked.
