---
title: "Promotion ROI Analysis"
author: "Tondreka Robles"
date: "Portfolio Version"
output:
  word_document: default
  pdf_document:
    number_sections: true
graphics: true
urlcolor: blue
---

\setlength{\parskip}{6pt}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 6, fig.height = 4.5, fig.align = "right")
```

## Part I: Promotional event planning

### Question 1

There is evidence of seasonal demand. Base cases are much higher for Events 1
and 2 than for the later events. Event 1 has 728 base cases and Event 2 has
1,360 base cases, while Events 3, 4, and 5 have 239, 449, and 333 base cases.
This pattern suggests that the earlier promotional periods occur during stronger
seasonal demand periods for the baking product.

### Question 2

Event 2 has the greater incremental sales response. Event 1's incremental
response is approximately 55 percent of base cases:

$$401 / 728 = 55.1\%$$

Event 2's incremental response is approximately 69 percent of base cases:

$$944 / 1360 = 69.4\%$$

Therefore, Event 2 produces the stronger incremental response relative to base
sales.

### Question 3

Events 2 and 5 are the profitable events. Event 5 has the highest ROI at about
53 percent, followed by Event 2 at about 44 percent. Event 1 is slightly
unprofitable with an ROI of about -2 percent, and Events 3 and 4 are clearly
unprofitable with ROIs of about -22 percent and -79 percent.

The main lesson is that a promotion is not profitable just because it creates
incremental volume. The event must generate enough incremental contribution to
cover variable costs, MDF/fixed payments, and forward-buying costs. Event 5 is
especially attractive because its incremental contribution is high relative to
its event cost. Event 2 is also profitable because it combines strong baseline
demand with a strong incremental response. Events 3 and 4 show that high costs
or weak incremental response can make a promotion unattractive even when units
increase.

### Question 4

The recalculation for Events 1 and 2 under four weeks of forward buying is shown
in the new workbook sheet named `Q4_Forward_Buy`. Increasing forward buying from
two weeks to four weeks doubles the forward-buying cases from 229 to 458 for
Events 1 and 2. That increases forward-buying cost from $962 to $1,924.

With four weeks of forward buying, Event 1 becomes more negative: event cost
increases to $9,164, event gross contribution falls to -$1,145, and ROI falls
to about -12.5 percent. Event 2 remains profitable but weaker: event cost rises
to $14,098, event gross contribution falls to $4,776, and ROI falls to about
33.9 percent. This shows that forward buying can materially reduce promotion
profitability because it increases manufacturer cost without increasing
consumer demand.

### Question 5

The Booz Allen Hamilton approach is more conservative because it includes more
promotion-related costs in the denominator. The class method is easier to
interpret when the goal is to isolate incremental profitability from the
promotion. I would use the class method for comparing event effectiveness and
the Booz Allen approach as a conservative financial check. Together, they show
both demand impact and cost exposure.

## Part II: Estimating lift factors and promotion ROI analysis for Hellman's Mayo

```{r load-data}
load("Hellmans.RData")

hellmans_DF$price <- hellmans_DF$sales_dollars / hellmans_DF$sales_units
hellmans_DF$feature_pctacv <- hellmans_DF$feature_pctacv / 100
hellmans_DF$display_pctacv <- hellmans_DF$display_pctacv / 100

dominicks <- subset(hellmans_DF, account == "Dominicks")
jewel <- subset(hellmans_DF, account == "Jewel")
```

### Question 1(c)

The feature and display variables are summarized separately by account below.

```{r q1c-summary}
summary(dominicks[, c("feature_pctacv", "display_pctacv")])
summary(jewel[, c("feature_pctacv", "display_pctacv")])
```

```{r q1c-histograms, fig.width=8, fig.height=6}
par(mfrow = c(2, 2))
hist(dominicks$feature_pctacv, main = "Dominick's Feature",
     xlab = "Feature %ACV", col = "lightblue")
hist(dominicks$display_pctacv, main = "Dominick's Display",
     xlab = "Display %ACV", col = "lightgreen")
hist(jewel$feature_pctacv, main = "Jewel Feature",
     xlab = "Feature %ACV", col = "lightblue")
hist(jewel$display_pctacv, main = "Jewel Display",
     xlab = "Display %ACV", col = "lightgreen")
par(mfrow = c(1, 1))
```

Feature is mostly zero and behaves more like an on/off promotional support
variable. When feature occurs, it often appears as a full feature event rather
than a smooth continuous variable. Display is more continuous, with a wider
range of values between zero and high display support. Jewel has more display
activity on average than Dominick's, and both accounts show that display varies
more continuously than feature.

### Question 1(d)

```{r q1d-correlations}
cor(dominicks[, c("display_pctacv", "feature_pctacv", "price")])
cor(jewel[, c("display_pctacv", "feature_pctacv", "price")])
```

Feature and display are positively correlated in both accounts, which suggests
that promotional tools are often coordinated. Price is negatively correlated
with feature and display because promotions usually combine support activity
with lower prices.

### Question 2

The regressions are estimated separately for Dominick's and Jewel.

```{r q2-models}
dom_price_model <- lm(log(sales_units) ~ log(price), data = dominicks)
dom_full_model <- lm(log(sales_units) ~ log(price) + feature_pctacv + display_pctacv,
                     data = dominicks)

jewel_price_model <- lm(log(sales_units) ~ log(price), data = jewel)
jewel_full_model <- lm(log(sales_units) ~ log(price) + feature_pctacv + display_pctacv,
                       data = jewel)

summary(dom_price_model)
summary(dom_full_model)
summary(jewel_price_model)
summary(jewel_full_model)
```

For Dominick's, the price-only model estimates a strong negative price
elasticity. When feature and display are added, the price coefficient becomes
less negative because part of the sales response that looked like a price effect
was actually due to coordinated feature and display support. In the full
Dominick's model, price, feature, and display are statistically significant.

For Jewel, the price-only model also estimates a strong negative price
elasticity. In the full model, price and display are statistically significant,
but feature is not. This means the feature coefficient should be set to zero
when calculating lift factors for Jewel.

### Question 3

Lift factors are calculated separately by account using the account-specific
full regressions from Question 2. Coefficients that are not statistically
significant at the 5 percent level are set to zero.

The correct TPR effect uses the log price change:

$$\exp(\beta_{price} \times \log(0.85))$$

not `exp(-0.15 * beta)`.

```{r q3-lift-factors}
sig_coefs <- function(model) {
  model_sum <- summary(model)
  coefs <- coef(model)
  pvals <- model_sum$coefficients[, 4]
  coefs[pvals[names(coefs)] >= 0.05 & names(coefs) != "(Intercept)"] <- 0
  coefs
}

lift_table <- function(model) {
  coefs <- sig_coefs(model)
  price_effect <- coefs["log(price)"] * log(0.85)
  display_effect <- coefs["display_pctacv"] * 0.70
  feature_effect <- coefs["feature_pctacv"] * 1.00

  data.frame(
    Promotion = c("15% TPR", "15% TPR + 70% Display",
                  "15% TPR + 70% Display + 100% Feature"),
    Lift = c(
      exp(price_effect),
      exp(price_effect + display_effect),
      exp(price_effect + display_effect + feature_effect)
    )
  )
}

dom_lifts <- lift_table(dom_full_model)
jewel_lifts <- lift_table(jewel_full_model)

dom_lifts
jewel_lifts
```

Dominick's has lift factors of approximately 1.35 for TPR alone, 2.42 for TPR
plus display, and 3.22 for TPR plus display plus feature. Jewel has lift factors
of approximately 1.36 for TPR alone and 2.88 for the two display-supported
promotions. Jewel's feature coefficient is not statistically significant, so
the full feature/display promotion has the same lift as the display-only
promotion.

### Question 4

The workbook `Part2_HellmansROI_amended.xlsx` contains the ROI calculations for
all three promotions for Dominick's and Jewel in the `ROI_Q3_Q4` sheet. The
calculations use account-specific regressions, account-specific lift factors,
and account-specific baseline units.

```{r q4-roi-functions}
baseline_units <- function(model) {
  coefs <- coef(model)
  sigma2 <- summary(model)$sigma^2
  exp(coefs["(Intercept)"] + coefs["log(price)"] * log(1.20) + sigma2 / 2)
}

roi_table <- function(model, account_name, fixed_display, fixed_full,
                      forward_buy_pct = 0.50) {
  lifts <- lift_table(model)$Lift
  base_units <- as.numeric(baseline_units(model))
  base_price <- 1.20
  promoted_price <- base_price * 0.85
  tpr_allowance <- base_price - promoted_price
  regular_vcm <- 0.55
  promoted_margin <- regular_vcm - tpr_allowance
  fixed_costs <- c(0, fixed_display, fixed_full)

  total_units <- base_units * lifts
  incremental_units <- total_units - base_units
  forward_buy_units <- base_units * forward_buy_pct
  incremental_contribution <- promoted_margin * incremental_units
  variable_cost <- tpr_allowance * base_units
  forward_buy_cost <- tpr_allowance * forward_buy_units
  event_cost <- variable_cost + fixed_costs + forward_buy_cost
  gross_contribution <- incremental_contribution - event_cost
  roi <- 100 * gross_contribution / event_cost

  data.frame(
    Account = account_name,
    Promotion = lift_table(model)$Promotion,
    Baseline_Units = base_units,
    Lift = lifts,
    Incremental_Units = incremental_units,
    Event_Cost = event_cost,
    Gross_Contribution = gross_contribution,
    ROI = roi
  )
}

dom_roi <- roi_table(dom_full_model, "Dominick's", 3000, 4500, 0.50)
jewel_roi <- roi_table(jewel_full_model, "Jewel", 5000, 6800, 0.50)

dom_roi
jewel_roi
```

### Question 5

```{r q5-no-forward-buy}
dom_roi_no_fb <- roi_table(dom_full_model, "Dominick's", 3000, 4500, 0)
jewel_roi_no_fb <- roi_table(jewel_full_model, "Jewel", 5000, 6800, 0)

dom_roi_no_fb
jewel_roi_no_fb
```

The 15 percent TPR-only promotion has the lowest ROI because it creates lift but
does not generate enough incremental contribution relative to the cost of the
price reduction and forward buying. The display-supported promotion is stronger
because display creates a much larger sales response. At Dominick's, the full
feature/display promotion is strongest because feature is statistically
significant and adds lift. At Jewel, the full feature/display promotion is less
attractive because feature is not statistically significant, so the added fixed
cost is not offset by additional lift.

Eliminating forward buying improves ROI for all promotions because it removes a
cost that does not increase consumed units. However, it does not change the main
ranking of promotions. The biggest lesson is that promotion profitability
depends on both demand response and cost structure. Strong lift is valuable only
when the added contribution is large enough to cover TPR costs, MDF/fixed
payments, and forward-buying costs.
